Gut-resident microorganisms and their genes are associated with cognition and neuroanatomy in children

### 2023-03-31 manuscript/main.tex
vanja comments 4

Stool samples (n=493) were collected by parents in OMR-200 tubes
(OMNIgene GUT, DNA Genotek, Ottawa, Ontario, Canada), stored on ice, and
brought within 24 hrs to the lab in RI where they were immediately
frozen at -80$^{\circ}$C. Stool samples were not collected if the subject had
taken antibiotics within the last two weeks. DNA extraction was
performed at Wellesley College (Wellesley, MA). Nucleic acids were
extracted from stool samples using the RNeasy PowerMicrobiome kit,
excluding the DNA degradation steps. Briefly, the samples were lysed by
bead beating using the Powerlyzer 24 Homogenizer (Qiagen, Germantown,
MD) at 2500 rpm for 45 s and then transferred to the QIAcube (Qiagen,
Germantown, MD) to complete the extraction protocol. Extracted DNA was
sequenced at the Integrated Microbiome Resource (IMR, Dalhousie
University, NS, Canada).

@@

Prediction of cognitive scores was carried out as a set of regression
experiments targeting real-valued continuous assessment scores.
Different experiment sets were designed to probe how different
representations of the gut microbiome (taxonomic profiles, Functional
profiles encoded as ECs) would behave, with and without the addition of
demographics (sex and maternal education as a proxy of socioeconomic
status) on participants from different age groups. Age (in months) was
provided as a covariate for all models (Table S3).

Random Forests (RFs)
\cite{breimanRandomForests2001}
were selected as the prediction engine and processed using the
DecisionTree.jl \cite{sadeghiDecisionTreeJlJulia2022}
implementation, inside the MLJ.jl \cite{blaomMLJJuliaPackage2020}.

Machine Learning framework. Independent RFs were trained for each
experiment, using a set of default regression hyperparameters from
Breiman and Cutler \cite{breimanRandomForests2001},
on a repeated cross-validation approach with different RNG seeds. One hundred
repetitions of 3-fold CV with 10 different intra-fold RNG states each
were employed, for a total of 3000 experiments per input set.

@@

T1-weighted anatomical MRI data were acquired using a 3-Dimensional MP-RAGE sequence
on 3T Siemens Trio scanner with the following parameters:
TE=5.6msec, TR=1400msec, TI = 950msec, FA=15 degrees, and 1.1x1.1x1.1mm resolution.
Image matrix and field-of-view were adjusted by infant size to achieve the desired image resolution.
For children under ~4 years of age, MRI was performed during natural non-sedated sleep using previously described methods\cite{deanPediatricNeuroimagingUsing2014}.
Older children were scanned awake whilst watching a movie or favorite TV show.
To help minimize motion, pediatric (MedVac) immobilizers were used along with foam cushions and padding.
Image quality was monitored during and following scanning.
If motion artifacts were observed (e.g., ghosting, blurring, or other signal variations),
scans were repeated or removed from analysis.

Following data acquisition, brain region volumes corresponding to total white and gray matter,
cortical gray matter, brainstem, and left and right hemisphere lateral ventricles,
thalamus, caudate, putamen, pallidum, hippocampus, amygdala,
and nucleus accumbens were calculated using an atlas matching approach\cite{bruchhageLongitudinalBrainCognitive}.
Individual images were first nonlinearly aligned to MNI-space using
a multi-step registration approach (ANTs 2.2 toolbox, \cite{avantsInsightToolKitImage2014}).
The MNI-aligned Harvard-Oxford brain atlas was then aligned and superimposed onto the individual child
by applying the inverse transformation, and the brain regions of interest delinated (segmented)
and their volune calculated by summing the number of image voxels within each\cite{jenkinsonFSL2012}.

Brain region volumes used in RF models were normalized to total brain volume
(the sum of white and gray matter).

@@

All data needed to evaluate the conclusions in the paper
are present in the paper and/or the Supplementary Materials.
Taxonomic and functional microbial profiles, as well as subject
demographics necessary for statistical analyses and machine learning are
available on the Open Science Framework
\cite{bonhamECHORESONANCEMicrobiome2022}.
Data processing, generation of summary statistics, and
generation of plots was performed using the julia programming language
and 3rd party packages
\cite{bezansonJuliaFreshApproach2017,
    bonhamMicrobiomeJlBiobakeryUtils2021,
    danischMakieJlFlexible2021,
    blaomMLJJuliaPackage2020,
    ben_sadeghi_2022_7359268,
    dahua_lin_2023_7695673,
    douglas_bates_2023_7734970}.
Information for replicating the package environment
as well as code for data analysis and figure generation, as well as scripts for automated
download of input files are available on github \cite{kevin_bonham_2023_7647510}.

@@

\printbibliography

>>>>

Stool samples (n = 493) were collected by parents in OMR-200 tubes
(OMNIgene GUT, DNA Genotek, Ottawa, Ontario, Canada), stored on ice, and
brought within 24 hrs to the lab in RI where they were immediately
frozen at -80$^{\circ}$C. Stool samples were not collected if the subject had
taken antibiotics within the last two weeks. DNA extraction was
performed at Wellesley College (Wellesley, MA). Nucleic acids were
extracted from stool samples using the RNeasy PowerMicrobiome kit,
excluding the DNA degradation steps. Briefly, the samples were lysed by
bead beating using the Powerlyzer 24 Homogenizer (Qiagen, Germantown,
MD) at 2500 rpm for 45 s and then transferred to the QIAcube (Qiagen,
Germantown, MD) to complete the extraction protocol. Extracted DNA was
sequenced at the Integrated Microbiome Resource (IMR, Dalhousie
University, NS, Canada).

@@

Prediction of cognitive scores was carried out as a set of regression
experiments targeting real-valued continuous assessment scores.
Different experiment sets were designed to probe how different
representations of the gut microbiome (taxonomic profiles, functional
profiles encoded as ECs) would behave, with and without the addition of
demographics (sex and maternal education as a proxy of socioeconomic
status) on participants from different age groups. Age (in months) was
provided as a covariate for all models (Table S3).

Random Forests (RFs)
\cite{breimanRandomForests2001}
were selected as the prediction engine and processed using the
DecisionTree.jl \cite{sadeghiDecisionTreeJlJulia2022}
implementation, inside the MLJ.jl \cite{blaomMLJJuliaPackage2020}.
Independent RFs were trained for each
experiment, using a set of default regression hyperparameters from
Breiman and Cutler \cite{breimanRandomForests2001},
on a repeated cross-validation approach with different RNG seeds. One hundred
repetitions of 3-fold CV with 10 different intra-fold RNG states each
were employed, for a total of 3000 experiments per input set.

@@

T1-weighted anatomical MRI data were acquired using a 3-Dimensional MP-RAGE sequence
on 3T Siemens Trio scanner with the following parameters:
TE=5.6msec, TR=1400msec, TI = 950msec, FA=15 degrees, and 1.1x1.1x1.1mm resolution.
Image matrix and field-of-view were adjusted by infant size to achieve the desired image resolution.
For children under 4 years of age, MRI was performed during natural non-sedated sleep using previously described methods \cite{deanPediatricNeuroimagingUsing2014}.
Older children were scanned awake whilst watching a movie or favorite TV show.
To help minimize motion, pediatric (MedVac) immobilizers were used along with foam cushions and padding.
Image quality was monitored during and following scanning.
If motion artifacts were observed (e.g., ghosting, blurring, or other signal variations),
scans were repeated or removed from analysis.

Following data acquisition, brain region volumes corresponding to total white and gray matter,
cortical gray matter, brainstem, and left and right hemisphere lateral ventricles,
thalamus, caudate, putamen, pallidum, hippocampus, amygdala,
and nucleus accumbens were calculated using an atlas matching approach \cite{bruchhageLongitudinalBrainCognitive}.
Individual images were first nonlinearly aligned to MNI-space using
a multi-step registration approach (ANTs 2.2 toolbox \cite{avantsInsightToolKitImage2014}).
The MNI-aligned Harvard-Oxford brain atlas was then aligned and superimposed onto the individual child
by applying the inverse transformation, and the brain regions of interest delinated (segmented)
and their volume calculated by summing the number of image voxels within each \cite{jenkinsonFSL2012}.
Brain region volumes used in RF models were normalized to total brain volume
(the sum of white and gray matter).

@@

All data needed to evaluate the conclusions in the paper
are present in the paper and/or the Supplementary Materials.
Taxonomic and functional microbial profiles, as well as subject
demographics necessary for statistical analyses and machine learning are
available on the Open Science Framework
\cite{bonhamECHORESONANCEMicrobiome2022}.
Data processing, generation of summary statistics, and
generation of plots was performed using the julia programming language
and 3rd party packages
\cite{bezansonJuliaFreshApproach2017,
    bonhamMicrobiomeJlBiobakeryUtils2021,
    danischMakieJlFlexible2021,
    blaomMLJJuliaPackage2020,
    ben_sadeghi_2022_7359268,
    dahua_lin_2023_7695673,
    douglas_bates_2023_7734970}.
Information for replicating the package environment
as well as code for data analysis and figure generation, as well as scripts for automated
download of input files are available on GitHub \cite{kevin_bonham_2023_7647510}.

@@

\printbibliography[heading=none]
